Arriaga meets Kitagawa. life expectancy decompositions including population subgroups

### 2024-01-25 manuscript/manuscript.Rmd
late updates; time to find an empirical example

# Introduction

@@

Sometimes we have a situation where we would like to decompose a difference between two populations that are each composed of like-defined subpopulations. For example, life expectancy in France versus Spain, each with education-specific subpopulations. In such situations, a decomposition should tell us the contribution to the difference in overall life expectancy due to rate differences in each age in each subpopulation, and also separate an effect due to overall compositional change. In this paper, I would like to propose a straightforward analytic method to decompose in this way. In short, within-subpopulation life expectancy differences can be handled by the Arriaga method. The case of weighting together the life expectancies of subpopulations by way of a composed radix can be handled using the well-known Kitagawa method. The elements of the value component of the Kitagawa method tell us how to rescale the Arriaga results specific to each subpopulation so as to isolate the age-subpopulation-specific effects on the overall life expectancy difference. I will justify this rescaling, and show that the results of this procedure are fully consistent with the results of a @horiuchi2008decomposition reframing of the problem.

@@

\begin{equation}
\label{eq_avg}
e_0 = \sum_s^S \pi^s \cdot e_0^s \quad \textrm{,}
\end{equation}

\noindent where the $S$ total subgroup fractions comprise a mixing distribution, $1 = \sum_s^S \pi^s$.

@@

\begin{equation}
\label{eq_kit_comp}
\Delta^\text{composition} = \sum_s^S \left(\pi^{s,t2}- \pi^{s,t1}\right) \cdot \frac{e_0^{s,t1} + e_0^{s,t2}}{2} \quad \textrm{,}
\end{equation}

@@

\begin{equation}
\label{eq_kit_sum}
\Delta = \Delta^\text{composition}  + \sum_s^S \Delta^{e_0^s\text{~change}}
\end{equation}

@@

\begin{align}
\label{eq_kit_e0}
\Delta^s &= e_0^{s,2} - e_0^{s,1}  \\
 &= \sum _x^\omega {}_n\Delta_x^s \quad \textrm{,}\nonumber 
\end{align}

\noindent where $\omega$ is the highest (open) age group, and $\Delta^s$ is the subgroup-specific ($s$) difference in life expectancy being decomposed, and it is composed of age-specific contributions, $\overrightarrow{{}_n\Delta_x^s}$ or $\overleftarrow{{}_n\Delta_x^s}$ depending on whether we decompose forward in time or backward, respectively (Riffe et. al., 2024). The forward age-specific values $\overrightarrow{{}_n\Delta_x^s}$ can be calculated following Arriaga's decomposition method following eq \eqref{eq_arriaga1}, consistent with @arriaga1984measuring or the presentation in @preston2000demography. I use a lifetable radix of 1, meaning $\ell_0 = 1$, to reduce the formula slightly.

@@

\begin{align}
\label{eq_sum2}
 -\Delta^s &= \sum_x^\omega \overleftarrow{{}_n\Delta_x^s} \\
 &= e_0^1 - e_0^2 \nonumber
\end{align}

@@

The quantity ${}_n\Delta_x^s$ given in eq \eqref{eq_arriaga3} is on the scale of group $s$ specific life expectancy changes, but to know its net impact on overall life expectancy change we rescale per eq \eqref{eq_rescale} to match the effect sizes from eq \eqref{eq_kit_e0}

\begin{equation}
\label{eq_rescale}
{}_n\Delta_x^{s,\text{net effect}} = \Delta^{e_0^s\text{~change}} \cdot \frac{{}_n\Delta_x^s}{\Delta^s}
\end{equation}

@@

Acknowledgments are not compulsory. Where included they should be brief. Grant or contribution numbers may be acknowledged.

Please refer to Journal-level guidance for any specific requirements.

@@

Some journals require declarations to be submitted in a standardised format. Please check the Instructions for Authors of the journal to which you are submitting to see if you need to complete this section. If yes, your manuscript must contain the following sections under the heading \`Declarations':

-   Funding
-   Conflict of interest/Competing interests (check journal-specific guidelines for which heading to use)
-   Ethics approval
-   Consent to participate
-   Consent for publication
-   Availability of data and materials
-   Code availability
-   Authors' contributions

\noindent If any of the sections are not relevant to your manuscript, please include the heading and write \`Not applicable' for that section.

>>>>

# Background

@@

# Introduction
A demographer might like to decompose a difference in life expectancy between two populations (or one population at two times) that are each composed of like-defined subpopulations. For example, life expectancy in France versus Spain, each with education-specific subpopulations, or a national life expectancy calculated as the weighted average of its regions at two points in time. In such situations, a decomposition would ideally tell us the contribution to the difference in overall life expectancy due to rate differences in each age in each subpopulation, and also separate an effect due to overall compositional change. In this paper, I would like to propose a straightforward analytic method to decompose in this way. In short, within-subpopulation life expectancy differences can be handled by the Arriaga method. The case of weighting together the life expectancies of subpopulations by way of a composed radix can be handled using the well-known Kitagawa method. The elements of the value component of the Kitagawa method tell us how to rescale the Arriaga results specific to each subpopulation so as to isolate the age-subpopulation-specific effects on the overall life expectancy difference. I will justify this rescaling, and show that the results of this procedure are fully consistent with the results of a @horiuchi2008decomposition reframing of the problem.

@@

\begin{equation}
\label{eq_avg}
e_0 = \sum_s \pi^s \cdot e_0^s \quad \textrm{,}
\end{equation}

\noindent where the $S$ total subgroup fractions comprise a mixing distribution, $1 = \sum_s \pi^s$.

@@

\begin{equation}
\label{eq_kit_comp}
\Delta^\text{composition} = \sum_s \left(\pi^{s,t2}- \pi^{s,t1}\right) \cdot \frac{e_0^{s,t1} + e_0^{s,t2}}{2} \quad \textrm{,}
\end{equation}

@@

\begin{equation}
\label{eq_kit_sum}
\Delta = \Delta^\text{composition}  + \sum_s \Delta^{e_0^s\text{~change}}
\end{equation}

@@

\begin{align}
\label{eq_kit_e0}
\Delta^s &= e_0^{s,2} - e_0^{s,1}  \\
 &= \sum _x {}_n\Delta_x^s \quad \textrm{,}\nonumber 
\end{align}

\noindent where $\Delta^s$ is the subgroup-specific ($s$) difference in life expectancy being decomposed, and it is composed of age-specific contributions, $\overrightarrow{{}_n\Delta_x^s}$ or $\overleftarrow{{}_n\Delta_x^s}$ depending on whether we decompose forward in time or backward, respectively (Riffe et. al., 2024). The forward age-specific values $\overrightarrow{{}_n\Delta_x^s}$ can be calculated following Arriaga's decomposition method following eq \eqref{eq_arriaga1}, consistent with @arriaga1984measuring or the presentation in @preston2000demography. I use a lifetable radix of 1, meaning $\ell_0 = 1$, to reduce the formula slightly.

@@

\begin{align}
\label{eq_sum2}
 -\Delta^s &= \sum_x \overleftarrow{{}_n\Delta_x^s} \\
 &= e_0^1 - e_0^2 \nonumber
\end{align}

@@

The age-specific quantities ${}_n\Delta_x^s$ derived from eq \eqref{eq_arriaga3} are on the scale of subgroup-specific life expectancy changes. To know the net impact of ${}_n\Delta_x^s$ on overall life expectancy change we rescale per eq \eqref{eq_rescale} to match the effect sizes from eq \eqref{eq_kit_e0}

\begin{equation}
\label{eq_rescale}
{}_n\Delta_x^{s,\text{net}} = \Delta^{e_0^s\text{~change}} \cdot \frac{{}_n\Delta_x^s}{\Delta^s} \quad \mathrm{.}
\end{equation}

With eqs \eqref{eq_kit_comp} and \eqref{eq_rescale}, we have all elements of the present decomposition:

\begin{equation}
\label{eq_full}
\Delta = \Delta^\text{composition} + \sum_s \sum_x {}_n\Delta_x^{s,\text{net}}
\end{equation}

@@

Thanks to Michael Lachanski for posing to me the question that led to the solution I present here, to Jonas Schoeley for first bringing the composition problem with the Kitagawa method to my attention, and to the MPIDR Laboratory of Population Health for posing assorted questions in September 2023 that got me thinking in these lines.

@@

Some journals require declarations to be submitted in a standardised format. Please check the Instructions for Authors of the journal to which you are submitting to see if you need to complete this section. If yes, your manuscript must contain the following sections under the heading

## Funding
This work was financed in part by grant nr SR22-00502 from _la Caixa Foundation_ and grant nr PID2022-142762OA-I00 from the Spanish _Ministerio de Ciencia e Innovación_.

## Conflict of interest
The author declares no conflict of interest

## Availability of data and materials
All data are ___

## Code availability
All code used here is available in an open `GitHub` repository \url{https://github.com/timriffe/arriagakitagawa}

### 2024-01-25 manuscript/manuscript.tex
late updates; time to find an empirical example

@@

Sometimes we have a situation where we would like to decompose a
difference between two populations that are each composed of
like-defined subpopulations. For example, life expectancy in France
versus Spain, each with education-specific subpopulations. In such
situations, a decomposition should tell us the contribution to the
difference in overall life expectancy due to rate differences in each
age in each subpopulation, and also separate an effect due to overall
compositional change. In this paper, I would like to propose a
straightforward analytic method to decompose in this way. In short,
within-subpopulation life expectancy differences can be handled by the
Arriaga method. The case of weighting together the life expectancies of
subpopulations by way of a composed radix can be handled using the
well-known Kitagawa method. The elements of the value component of the
Kitagawa method tell us how to rescale the Arriaga results specific to
each subpopulation so as to isolate the age-subpopulation-specific
effects on the overall life expectancy difference. I will justify this
rescaling, and show that the results of this procedure are fully
consistent with the results of a \citet{horiuchi2008decomposition}
reframing of the problem.

@@

\begin{equation}
\label{eq_avg}
e_0 = \sum_s^S \pi^s \cdot e_0^s \quad \textrm{,}
\end{equation}

\noindent where the \(S\) total subgroup fractions comprise a mixing
distribution, \(1 = \sum_s^S \pi^s\).

@@

\begin{align}
\label{eq_kit_e0}
\Delta^s &= e_0^{s,2} - e_0^{s,1}  \\
 &= \sum _x^\omega {}_n\Delta_x^s \quad \textrm{,}\nonumber 
\end{align}

\noindent where \(\omega\) is the highest (open) age group, and
\(\Delta^s\) is the subgroup-specific (\(s\)) difference in life
expectancy being decomposed, and it is composed of age-specific
contributions, \(\overrightarrow{{}_n\Delta_x^s}\) or
\(\overleftarrow{{}_n\Delta_x^s}\) depending on whether we decompose
forward in time or backward, respectively (Riffe et. al., 2024). The
forward age-specific values \(\overrightarrow{{}_n\Delta_x^s}\) can be
calculated following Arriaga's decomposition method following eq
\eqref{eq_arriaga1}, consistent with \citet{arriaga1984measuring} or the
presentation in \citet{preston2000demography}. I use a lifetable radix
of 1, meaning \(\ell_0 = 1\), to reduce the formula slightly.

@@

The quantity \({}_n\Delta_x^s\) given in eq \eqref{eq_arriaga3} is on
the scale of group \(s\) specific life expectancy changes, but to know
its net impact on overall life expectancy change we rescale per eq
\eqref{eq_rescale} to match the effect sizes from eq \eqref{eq_kit_e0}

\begin{equation}
\label{eq_rescale}
{}_n\Delta_x^{s,\text{net effect}} = \Delta^{e_0^s\text{~change}} \cdot \frac{{}_n\Delta_x^s}{\Delta^s}
\end{equation}

@@

Some journals require declarations to be submitted in a standardised
format. Please check the Instructions for Authors of the journal to
which you are submitting to see if you need to complete this section. If
yes, your manuscript must contain the following sections under the
heading `Declarations':

\begin{itemize}
\tightlist
\item
  Funding
\item
  Conflict of interest/Competing interests (check journal-specific
  guidelines for which heading to use)
\item
  Ethics approval
\item
  Consent to participate
\item
  Consent for publication
\item
  Availability of data and materials
\item
  Code availability
\item
  Authors' contributions
\end{itemize}

\noindent If any of the sections are not relevant to your manuscript,
please include the heading and write `Not applicable' for that section.

>>>>

\hypertarget{background}{%
\section{Background}\label{background}}

@@

A demographer might like to decompose a difference in life expectancy
between two populations (or one population at two times) that are each
composed of like-defined subpopulations. For example, life expectancy in
France versus Spain, each with education-specific subpopulations, or a
national life expectancy calculated as the weighted average of its
regions at two points in time. In such situations, a decomposition would
ideally tell us the contribution to the difference in overall life
expectancy due to rate differences in each age in each subpopulation,
and also separate an effect due to overall compositional change. In this
paper, I would like to propose a straightforward analytic method to
decompose in this way. In short, within-subpopulation life expectancy
differences can be handled by the Arriaga method. The case of weighting
together the life expectancies of subpopulations by way of a composed
radix can be handled using the well-known Kitagawa method. The elements
of the value component of the Kitagawa method tell us how to rescale the
Arriaga results specific to each subpopulation so as to isolate the
age-subpopulation-specific effects on the overall life expectancy
difference. I will justify this rescaling, and show that the results of
this procedure are fully consistent with the results of a
\citet{horiuchi2008decomposition} reframing of the problem.

@@

\begin{equation}
\label{eq_avg}
e_0 = \sum_s \pi^s \cdot e_0^s \quad \textrm{,}
\end{equation}

\noindent where the \(S\) total subgroup fractions comprise a mixing
distribution, \(1 = \sum_s \pi^s\).

@@

\begin{align}
\label{eq_kit_e0}
\Delta^s &= e_0^{s,2} - e_0^{s,1}  \\
 &= \sum _x {}_n\Delta_x^s \quad \textrm{,}\nonumber 
\end{align}

\noindent where \(\Delta^s\) is the subgroup-specific (\(s\)) difference
in life expectancy being decomposed, and it is composed of age-specific
contributions, \(\overrightarrow{{}_n\Delta_x^s}\) or
\(\overleftarrow{{}_n\Delta_x^s}\) depending on whether we decompose
forward in time or backward, respectively (Riffe et. al., 2024). The
forward age-specific values \(\overrightarrow{{}_n\Delta_x^s}\) can be
calculated following Arriaga's decomposition method following eq
\eqref{eq_arriaga1}, consistent with \citet{arriaga1984measuring} or the
presentation in \citet{preston2000demography}. I use a lifetable radix
of 1, meaning \(\ell_0 = 1\), to reduce the formula slightly.

@@

The age-specific quantities \({}_n\Delta_x^s\) derived from eq
\eqref{eq_arriaga3} are on the scale of subgroup-specific life
expectancy changes. To know the net impact of \({}_n\Delta_x^s\) on
overall life expectancy change we rescale per eq \eqref{eq_rescale} to
match the effect sizes from eq \eqref{eq_kit_e0}

\begin{equation}
\label{eq_rescale}
{}_n\Delta_x^{s,\text{net}} = \Delta^{e_0^s\text{~change}} \cdot \frac{{}_n\Delta_x^s}{\Delta^s} \quad \mathrm{.}
\end{equation}

With eqs \eqref{eq_kit_comp} and \eqref{eq_rescale}, we have all
elements of the present decomposition:

\begin{equation}
\label{eq_full}
\Delta = \Delta^\text{composition} + \sum_s \sum_x {}_n\Delta_x^{s,\text{net}}
\end{equation}

@@

Some journals require declarations to be submitted in a standardised
format. Please check the Instructions for Authors of the journal to
which you are submitting to see if you need to complete this section. If
yes, your manuscript must contain the following sections under the
heading

\hypertarget{funding}{%
\subsection{Funding}\label{funding}}

This work was financed in part by grant nr SR22-00502 from \emph{la
Caixa Foundation} and grant nr PID2022-142762OA-I00 from the Spanish
\emph{Ministerio de Ciencia e Innovación}.

\hypertarget{conflict-of-interest}{%
\subsection{Conflict of interest}\label{conflict-of-interest}}

The author declares no conflict of interest

\hypertarget{availability-of-data-and-materials}{%
\subsection{Availability of data and
materials}\label{availability-of-data-and-materials}}

All data are \_\_\_

\hypertarget{code-availability}{%
\subsection{Code availability}\label{code-availability}}

All code used here is available in an open \texttt{GitHub} repository
\url{https://github.com/timriffe/arriagakitagawa}
